---
title: "Population graphs and f statistics"
author: "Margaux Lefebvre"
date: "`r Sys.Date()`"
output:
  github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE) #prevent all chunks from running
library(tidyverse)
```

# Nucleotide diversity (pi) and Tajima's D

## Make tandardized sample size

Version: R v4.2.

To not have a biased estimators, it's better to have minimum 10 samples ([paper here](https://doi.org/10.1002/ece3.1846)).

```{r}
bams_infos <- read_delim("./Data/bams_infos.txt", 
    delim = "\t", escape_double = FALSE, 
    col_names = c("path","sample_ID"), trim_ws = TRUE)

# add meta information
Samples_data<-read_delim("./Data/metadata_vivax.tsv", 
     delim = "\t", show_col_types = FALSE)
Samples_data<-Samples_data[,-c(3:8,10:15,17)]
colnames(Samples_data)<-c("sample_ID","species","country","population","cluster")
info<-inner_join(bams_infos,Samples_data)

# Which populations do we keep ?
min_size<-10

number_data<-info %>% 
  group_by(country) %>%
  summarise(N = length(country))

kept_populations<-subset(number_data, number_data$N>=min_size)
kept_populations

# Make the file for each pop, keeping sample at random
for (pop in kept_populations$country){
  temp<-subset(info, info$country==pop)
  samples_kept<-temp[sample(nrow(temp), min_size), ]
  write_tsv(as.data.frame(samples_kept$path), file = paste0("./Data/list_bams_",pop,".10.stand.txt"), col_names = F)
}
```

# Calculation of pi and Tajima's D

Version: ANGSD v0.940.

```{bash}
# Make ancestral fasta with Plasmodium vivax-like
angsd -b PL.list.txt -out PL.ancestral \
        -minMapQ 20 -minQ 20 -setMaxDepthInd 106 -doCounts 1 \
        -doFasta 2 -nThreads 12
        
echo "--> Start processing: $POP" #Calculation for each population

# Create Beagle output
angsd -b ./diversity/list_bams_$POP.10.stand.txt -rf core_regions.txt \
        -minMapQ 20 -minQ 20 -minInd 5 \
        -GL 2 -nThreads 12 -doSaf 1 -anc PL.ancestral.fa.gz -out ./diversity/$POP.10


#Calculate pi and Tajima's D
./angsd-0.940/bin/realSFS ./diversity/$POP.10.saf.idx -P 12 > ./diversity/$POP.10.sfs

./angsd-0.940/bin/realSFS saf2theta ./diversity/$POP.10.saf.idx -sfs ./diversity/$POP.10.sfs -outname ./diversity/$POP.10

./angsd-0.940/bin/thetaStat do_stat ./diversity/$POP.10.thetas.idx -win 500 -step 500 -outnames ./diversity/$POP.10.thethasWindow500.gz # Sliding windows of 500pb, not overlapping
```

# Test for significant difference

Version: R v4.2.

```{r}
library(qqplotr)
library(MASS)
library(rstatix)
library(tidyverse)

# Read the data
pop <- "Afghanistan"
path<-paste0("./Data/",pop,".thethasWindow500.gz.pestPG")
data_diversity <- read_table(path)
data_diversity$cluster<-pop

POP=c("Mauritania", "Ethiopia","Sudan", "Malaysia", "Cambodia", "Vietnam", "Thailand", "Colombia", "Peru","Brazil","French_Guiana","Venezuela","India","Pakistan")
for (pop in POP){
  path<-paste0("./Data/",pop,".thethasWindow500.gz.pestPG")
temp <- read_table(path)
temp$cluster<-pop
  data_diversity<-rbind(data_diversity,temp)
}

data_diversity$pi<-data_diversity$tP/data_diversity$nSites # tP can be used to estimate the window-based pairwise nucleotide diversity (π), when we divide tP by the number of sites within the corresponding window (-nSites).

### Pi
# QQplot to check if distribution is normal
data_diversity %>%
  ggplot(aes(sample = pi)) +
     stat_qq_band() +
 stat_qq_line() +
 stat_qq_point()+ labs(x = "Theoretical Quantiles", y = "Sample Quantiles")+ ggtitle("QQ plot for \u03c0 values")+
  facet_wrap(~ cluster, scales = "free_y")+theme_bw()

# Test difference
knitr::kable(data_diversity %>% 
  wilcox_test(pi ~ cluster, p.adjust.method = "bonferroni"))
data_diversity %>% kruskal_test(pi ~ cluster)

### Tajima's D
# QQplot to check if distribution is normal
data_diversity %>%
  ggplot(aes(sample = Tajima)) +
     stat_qq_band() +
 stat_qq_line() +
 stat_qq_point()+ labs(x = "Theoretical Quantiles", y = "Sample Quantiles")+ ggtitle("QQ plot for Tajima's D values")+
  facet_wrap(~ cluster, scales = "free_y")+theme_bw()
# Test difference
knitr::kable(data_diversity %>% 
  wilcox_test(Tajima ~ cluster, p.adjust.method = "bonferroni"))
data_diversity%>% 
  group_by(cluster) %>%
  get_summary_stats(Tajima, type = c("quantile"))
```

# Coalescence with Relate and Colate

## Preparing the input files

Version: bcftools v1.10.2, ANGSD v0.940, R v4.2, vcftools v0.1.16, Relate v1.1.9.

First, I have to separate each chromosome, "phasing", and create hap/sample/legend files.

```{bash}
echo "--> Start processing chromosome: $CHR" #Deal with each chromosome separated

# Calling with ANGSD to have all the sites (only in the core genome), and only P. vivax samples
angsd -b list_bams.noPL.txt -ref P.vivax-PvP01_reference_genome.fasta -out temp_$CHR \
-r core_${CHR}.txt \
-minMapQ 20 -minQ 20 -minInd 5 -setMinDepth 5 -setMaxDepthInd 106 -doCounts 1 \
-nThreads 4 -gl 1 -dopost 1 -domajorminor 1 -domaf 1 -dobcf 1 --ignore-RG 0 -dogeno 1

# "Phasing the data" but it's artificial since it's haploid
bcftools view temp_$CHR.bcf.gz \
--output-type z --output-file vivax_${CHR}.vcf.gz
tabix vivax_${CHR}.vcf.gz

zcat vivax_${CHR}.vcf.gz | grep '^#' > vivax_${CHR}.vcf.header
zcat vivax_${CHR}.vcf.gz | grep -v '^#' > vivax_${CHR}.vcf.body

sed -e "s/\//|/g" vivax_${CHR}.vcf.body > out.vivax_${CHR}.vcf.body

# combine the new bcf body with the old header
cat vivax_${CHR}.vcf.header out.vivax_${CHR}.vcf.body | bgzip -c > vivax_${CHR}.phase.vcf.gz
rm vivax_${CHR}.vcf.* # remove intermediate files
rm  out.vivax_${CHR}.vcf.body

# Make the hap/sample/legend files
bcftools convert vivax_${CHR}.phase.vcf.gz --haplegendsample vivax_hap_${CHR}

#Calculate the proportion of missing data for each sample
vcftools --gzvcf vivax_${CHR}.vcf.gz --missing-indv --out sample_${CHR} 
```

Add the proportion of missing data to the sample file
```{r}
ARR_CHR=c("01","02","03","04","05","06","07","08","09","10","11","12","13","14")
for (chr in ARR_CHR){
  ind_miss  <- read_delim(paste0("./Data/sample_",chr,".imiss"), delim = "\t",
                        col_names = c("ID", "ndata", "nfiltered", "nmiss", "fmiss"), skip = 1)
ID_1<-c(ind_miss$ID)
ID_2<-c(ind_miss$ID)
missing<-c(ind_miss$fmiss)
temp_sample<-cbind(ID_1, ID_2, missing)
write.table(temp_sample, paste0("./Data/vivax_hap_",chr,".sample"), quote = FALSE, row.names = FALSE)
}
```

Convert from hap/legend/sample to haps/sample
```{bash}
echo "--> Start processing chromosome: $CHR" #Deal with each chromosome separated

RelateFileFormats --mode ConvertFromHapLegendSample --chr ${CHR} \
                 --haps vivax_${CHR}.haps \
                 --sample sample_${CHR}.sample \
                 -i vivax_hap_${CHR}
```

File for the outgroup to polarize the samples:
```{bash}
echo "--> Start processing chromosome: $CHR" #Deal with each chromosome separated

angsd -b PL.list.txt -out PL_${CHR}.ancestral \
        -r PvP01_${CHR}_v1 \
        -minMapQ 20 -minQ 20 -setMaxDepthInd 106 -doCounts 1 \
        -doFasta 2
```

The mask file is created based on the coregenome as defined by [Daron et *al.* (2017)](0.1126/sciadv.abc3713).

Create pop labels file (space delimited), with 4 columns: ID POP GROUP SEX
```{r}
library(readr)
vivax_samples <- read_table("./Data/haplegsam/sample_01.sample")
vivax_samples <-vivax_samples[,1]
colnames(vivax_samples)<-"ID"

lis_bams<-read_csv("./Data/list_samples_order.txt", 
    col_names = "ID")
## Add meta data
Samples_data<-read_delim("./Data/metadata_vivax.tsv", 
     delim = "\t", show_col_types = FALSE)
Samples_data<-Samples_data[,c(1,9,16)]
colnames(Samples_data)<-c("ID","POP","GROUP")
poplabels<-inner_join(lis_bams,Samples_data)
poplabels$POP <- sub(" ", "_", poplabels$POP)
poplabels$GROUP <- sub(" ", "_", poplabels$GROUP)
poplabels$GROUP <- sub(" ", "_", poplabels$GROUP)
poplabels$SEX <- NA
poplabels$ID<-vivax_samples$ID_1[-1]

write.table(poplabels, paste0("./Data/vivax.poplabels"), quote = FALSE, row.names = FALSE)
```

Prepare the input files
```{bash}
echo "--> Start processing chromosome: $CHR" #Deal with each chromosome separated

./scripts/PrepareInputFiles/PrepareInputFiles.sh \
                 --haps vivax_${CHR}.haps \
                 --sample sample_${CHR}.sample \
                 --ancestor PL_${CHR}.ancestral.fa \
                 --mask mask_${CHR}.fa \
                 --poplabels vivax.poplabels \
                 -o ./vivax_${CHR}_input
                 
./bin/Relate \
      --mode All -m 6.43333e-9 -N 20000 \
      --haps "./vivax_${CHR}_input.haps.gz" \
      --sample "./vivax_${CHR}_input.sample.gz" \
      --map "./Genetic_map_vivax_${CHR}.txt" \
      --annot "./vivax_${CHR}_input.annot" \
      --dist "./vivax_${CHR}_input.dist.gz" \
      -o vivax_relate_${CHR}
```

## Infer coalescence rates

Version: Relate v1.1.9, R v4.2.

```{bash}
./scripts/EstimatePopulationSize/EstimatePopulationSize.sh \
-i "./vivax_relate" \
-o "est_size_iter" \
-m 6.43333e-9 \
--chr "./chromosome_list.txt" \
--poplabels "vivax.poplabels"
--bins 0,7,0.25 \
--years_per_gen 0.18 

${PATH_TO_RELATE}/bin/RelateCoalescentRate \
--mode EstimatePopulationSize \
-i "est_size_iter" \
-o "est_size_total" \
--bins 0,7,0.25 \
--years_per_gen 0.18 \
--chr "./chromosome_list.txt" \
--poplabels "vivax.poplabels"
```

Plot
```{r}
library(relater)
library(ggplot2)
library(tidyverse)
# Read the dataset
coal <- read.coal("est_size_joint_results.coal")
#haploid effective population size is the 1* inverse coalescence rate
coal$popsize <- 1/coal$haploid.coalescence.rate
#multiply epochs times by 0.18 to scale to years (assuming 0.18 years per generation)
coal$epoch.start <- 0.18 * coal$epoch.start


#We subset to within group coalescence rates, but feel free to look at cross-population rates too
coal <- subset(coal, group1 == group2)
kept_pop<-c("Afghanistan","Brazil","Colombia","Guyana","French_Guiana", "Ethiopia", "Honduras", "India", "Mauritania", "Mexico", "PNG", "Peru", "Ecuador", "Venezuela","Thailand","Malaysia","Pakistan","Cambodia","Vietnam")
coal<-coal[(coal$group1 %in% kept_pop),] # Keep principal clusters for the plots

# Populations gathered by cluster to plot by regions.
Cluster<- rep(NA, length(coal$group1))
Cluster[grep("Pakistan", coal$group1)] <- "Middle-East and South Asia"
Cluster[grep("Afghanistan", coal$group1)] <- "Middle-East and South Asia"
Cluster[grep("India", coal$group1)] <- "Middle-East and South Asia"
Cluster[grep("Pakistan", coal$group1)] <- "Middle-East and South Asia"
Cluster[grep("Brazil", coal$group1)] <- "Latin America"
Cluster[grep("Ecuador", coal$group1)] <- "Latin America"
Cluster[grep("Guyana", coal$group1)] <- "Latin America"
Cluster[grep("French_Guiana", coal$group1)] <- "Latin America"
Cluster[grep("Honduras", coal$group1)] <- "Latin America"
Cluster[grep("Colombia", coal$group1)] <- "Latin America"
Cluster[grep("Mexico", coal$group1)] <- "Latin America"
Cluster[grep("Peru", coal$group1)] <- "Latin America"
Cluster[grep("Venezuela", coal$group1)] <- "Latin America"
Cluster[grep("Cambodia", coal$group1)] <- "East Asia"
Cluster[grep("Malaysia", coal$group1)] <- "East Asia"
Cluster[grep("Thailand", coal$group1)] <- "East Asia"
Cluster[grep("Vietnam", coal$group1)] <- "East Asia"
Cluster[grep("Ethiopia", coal$group1)] <- "Africa"
Cluster[grep("Mauritania", coal$group1)] <- "Africa"
Cluster[grep("PNG", coal$group1)] <- "East Asia"
Cluster[grep("Spain", coal$group1)] <- "Europe"
coal<-cbind(coal, Cluster)

# Plot
coal %>%
  ggplot( aes(x=epoch.start, y=1/haploid.coalescence.rate)) +
    annotation_logticks(short = unit(1, "npc"), mid = unit(1, "npc"), long = unit(1,"npc"), color="grey",alpha = 0.2)  +
    geom_step( data=coal  %>% dplyr::select(-Cluster), aes(group=group1), color="black", size=0.5, alpha=0.2) +
    geom_step( aes(color=group2), size=1.2 )+theme_bw()+
 theme(panel.grid = element_blank(), panel.background = element_blank(),
      axis.text.x = element_text(size = 11),
      axis.text.y = element_text(size = 11),
      axis.title.y = element_text(size=15),
     axis.title.x = element_text(size=15),
     strip.text = element_text(size = 12))+
  scale_x_log10(labels=scales::comma, sec.axis = sec_axis(~./0.18, name = expression(Generation~ago~(log[10]~scale~with~5.5~generations/year)), labels=scales::comma))+
  scale_y_log10(labels=scales::comma)+
    facet_wrap(Cluster~., ncol = 1, strip.position = "top")+
  xlab(expression(Years~ago~(log[10]~ scale)))+
  ylab(expression(Ne~(log[10]~ scale)))+ coord_cartesian(xlim =c(10, 1e5), ylim = c(1e3, 10e4))+
  scale_color_manual(values = c("#F76E11","#0E4C92","#DE3163","#4682B4","#4D516D","#5C8D89","#22577E","#6F8FAF","#73C2FB","#FF0000","#FE7D6A","#4CBB17","#A4EBF3","#880808","royalblue","#9C297F","#C65D7B","navyblue","#DA70D6"))+
  scale_fill_manual(values = c("#F76E11","#0E4C92","#DE3163","#4682B4","#4D516D","#5C8D89","#22577E","#6F8FAF","#73C2FB","#FF0000","#FE7D6A","#4CBB17","#A4EBF3","#880808","royalblue","#9C297F","#C65D7B","navyblue","#DA70D6"))
```

# Colate

## Create input files

Version: Colate v0.1.3.

The input files are from Relate. First we have to crate a `mut` file with fixed column file.
```{bash}
# Ref files for modern populations
echo "--> Start processing chromosome: $CHR" #Deal with each chromosome separated

vivax_relate_${CHR}

${PATH_TO_BINARY}/Colate --mode preprocess_mut \
	--anc ./vivax_relate_${CHR}.anc \
	--mut ./vivax_relate_${CHR}.mut \
	--reference_bcf ./vivax_ref_chr${CHR}.bcf.gz \
	--ref_genome ./P.vivax-PvP01_reference_chr${CHR}.fasta \
	--anc_genome ./PL_${CHR}.ancestral.fa \
	--mask ./mask_${CHR}.fa \
	-o ./vivax_colate_input_ref_chr${CHR}.mut
gzip ./vivax_colate_input_ref_chr${CHR}.mut

```

Precompute Colate input files
```{bash}
# For Ebro
Colate \
	--mode make_tmp \
	--mut vivax_colate_input_ref \
	--target_bam ./Ebre_composite.PV.mem.dedup.rg.rescaled.chr.bam \
	--ref_genome ./P.vivax-PvP01_reference \
  --chr chromosome_list.txt \
	-o vivax_Spain_out

# For modern population
echo "--> Start processing chromosome: $POP" #Deal with each population we want to compare to Ebro

Colate \
	--mode make_tmp \
	--mut vivax_colate_input_ref \
	--target_bcf ./vivax_${POP} --target_mask ./mask \
	--ref_genome ./P.vivax-PvP01_reference \
  --chr chromosome_list.txt \
	-o vivax_${POP}_out
```

## Infer cross-coalescent rate

Version: Colate v0.1.3, R v4.2.

```{bash}
echo "--> Start processing chromosome: $POP" #Deal with each population we want to compare to Ebro

Colate --mode mut \
--mut vivax_colate_input_ref \
--target_bam Ebre_composite.PV.mem.dedup.rg.rescaled.chr.bam \
--reference_tmp vivax_${POP}_out.colate.in \
--bins 0,7,0.25 \
--years_per_gen 0.18 \
--chr  chromosome_list.txt \
--target_age 75 \
--reference_age 0 \
--num_bootstraps 100 \
-o Spain_${POP}
```

Plot
```{r}
library(relater)
library(dplyr)
library(ggplot2)
library(pammtools)
library(cowplot)

  # Cross-coalescent Spain-modern pop
df <- data.frame()
for(pop in c("Colombia","French_Guiana")){

  df <- bind_rows(df, read.coal(paste0("./Data/Spain_",pop,".coal")) %>% 
								group_by(epoch.start) %>% 
								summarize(mean = mean(haploid.coalescence.rate), 
													lower = quantile(haploid.coalescence.rate, prob = 0.025),																							 
													upper = quantile(haploid.coalescence.rate, prob = 0.975),	
													num_bootstrap = length(haploid.coalescence.rate),
													target = "Ebro",
													reference = pop))
}

  # Cross-coalescent intra modern pop
df <- subset(df, mean > 0)
df_ref <- data.frame()

for(pop in c("Colombia","French_Guiana")){
  df_ref <- bind_rows(df_ref, read.coal(paste0("./Data/",pop,"_",pop,".coal")) %>% 
								group_by(epoch.start) %>% 
								summarize(mean = mean(haploid.coalescence.rate), 
													lower = quantile(haploid.coalescence.rate, prob = 0.025),																							 
													upper = quantile(haploid.coalescence.rate, prob = 0.975),	
													num_bootstrap = length(haploid.coalescence.rate),
													target = pop1,
													reference = pop))
}

# Make the plots
for(pop in c("Colombia","French_Guiana")){
temp_ref<-subset(df_ref, reference==pop)
temp_ref<-subset(temp_ref, target==pop)
temp_target<-subset(df, reference==pop)
p<-ggplot(temp_target) +
				geom_step(aes(x = 0.18*epoch.start, y = 1/mean, colour = paste0(target,"-",reference)), lwd = 1.1) + 
				geom_stepribbon(aes(x = 0.18*epoch.start, ymin = 1/upper, ymax = 1/lower, fill = paste0(target,"-",reference)), alpha = 0.5) + 
				annotation_logticks(short = unit(1, "npc"), mid = unit(1, "npc"), long = unit(1,"npc"), color="grey",alpha = 0.2) +
			 theme(panel.grid = element_blank(), panel.background = element_blank(),
      axis.text.x = element_text(size = 11),
      axis.text.y = element_text(size = 11),
      axis.title.y = element_text(size=15),
     axis.title.x = element_text(size=15),
     strip.text = element_text(size = 12),
     legend.title=element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=1))+
  scale_x_log10(labels=scales::comma, sec.axis = sec_axis(~./0.18, name = expression(Generation~ago~(log[10]~scale~with~5.5~generations/year)), labels=scales::comma))+
  scale_y_log10(labels=scales::comma)+
				ylab("Inverse coalescence rates") +
      	xlab("years ago")+
  xlab(expression(Years~ago~(log[10]~ scale)))+
  ylab(expression(1/coalescence~rate~(log[10]~ scale)))+
  	geom_step(data=temp_ref, aes(x = 0.18*epoch.start, y = 1/mean, colour = paste0(reference)), lwd = 1.1)+ 
				geom_stepribbon(data=temp_ref,aes(x = 0.18*epoch.start, ymin = 1/upper, ymax = 1/lower, fill = paste0(reference)), alpha = 0.5)+
  coord_cartesian(xlim =c(10, 1e5), ylim = c(1e3, 1e5))+ geom_vline(xintercept = 75, linetype="dotted", 
                color = "black", size=1)
assign(paste0(paste0("plot_Spain_",pop)), p)
}

plot_Spain_Colombia+
  scale_color_manual(values = c("#4682B4", "#FFC300"))+
  scale_fill_manual(values = c("#4682B4", "#FFC300"))

plot_Spain_French_Guiana+
  scale_color_manual(values = c("#FFC300","#6F8FAF"))+
  scale_fill_manual(values = c("#FFC300","#6F8FAF"))
```